Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-01-27 paper.Rnw
Fixed unclosed macros

THe gold standard for neuroanatomical segmentation is manual delineation by an
expert human rater.  However, with the availability of increasingly large MRI
datasets, the time and expertise required for manual segmentation becomes
prohibitive \citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. This
effort is complicated by the fact that there is significant variation between
segmentation protocols with respect to specific anatomical boundaries of the
hippocampus \citep{Geuze2004} and this has led to efforts to create an unified
hippocampal segmentation protocol \citep{Jack2011,Boccardi2013,Boccardi2013a}.
In addition, there is controversy over the appropriate manual segmentation
protocol to use in a particular imaging study \citep{Nestor2012}. Thus, a
segmentation algorithm that can easily adapt to different manual segmentation
definitions would be of significant benefit to the neuroimaging community.

@@

Also of interest to this manuscript are the methods that attempt to define
hippocampal subfields using standard T1- or T2-weighted data, of which there are
few.  \citet{VanLeemput2009} demonstrate that the applicability of hippocampal
subfield segmentation in T1-weighted images by Bayesian techniques using Markov
random field shape priors learned from 10 manual segmentations.  This work,
available as part of the \freesurfer{} package, is limited as the segmentation
omits the tail of the hippocampus and the protocol has yet to be fully
validated.  \citet{Yushkevich2009} manually segment hippocampal subfields on
high-resolution (either $0.2mm^3$ isotropic or $0.2mm \times 0.3mm \times 0.2mm$
resolution voxels) T2-weighted MR images acquired from five post-mortem medial
temporal lobe samples. Then, using nonlinear registration guided by shape-based
models of the subfield segmentations, and manually derived hippocampus masks of
the target images, the authors demonstrate accurate parcellation of hippocampal
subfields in clinical 3T T1-weighted MRI volumes.  Using \multiatlas with bias
correction techniques, \citet{Yushkevich2010} demonstrate a semi-automated
method of subfield segmentation on in vivo focal T2-weighted MR acquisitions of
the temporal lobe. Manual input is only needed to mark mark divisions between
the head, body and tail of the hippocampus on target images.

@@

The following section describes experiments conducted to assess the
segmentation quality of the \mb{} algorithm:  
\begin{itemize}
\item Experiment 1 investigates the accuracy of \mb{} whole hippocampus
segmentation of aging and Alzhiemer's diseased subjects over a wide range of
parameter settings using a Monte Carlo cross-validation design. The results of
this experiment enable us to choose the parameter settings offering the best
performance for use in subsequent experiments.  
\item Experiment 2 consists of a similar cross-validation design to verify these
parameter settings on brain images of young, first episode psychosis
patients. 
\item Experiment 3 compares \mb{} whole hippocampus segmentations with those of
well-known automated methods on the entire \adnidataset image dataset consisting
of 
\item Experiment 4 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. 
\Sexpr{length(means.complete$RID)} brain images of subjects diagnosed as
cognitively normal, having mild cognitive impairment, or Alzheimer's disease.
\end{itemize}

@@

% In this experiment we conduct 10 rounds of \mb{} and \multiatlas{} segmentation of
% of 60 subjects using atlas and template library sizes varying from 1-9 and 1-20
% images respectively, two registration algorithms (\ants{} or \animal{}), and three
% label fusion techniques (unweighted, cross-correlation, and normalised mutual
% information weighted majority vote).  Computed segmentations are compared to
% \psegs using Dice's Similarity Coefficient (DSC).  In the Supplementary
% Materials we have replicated this experiment using the \snt semi-automated
% labels included with the ADNI dataset.

@@

\subsubsection{Materials and Methods: \expadni}
\paragraph{\adnidataset dataset} 
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
\snt{}, \maper, \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).

@@

The core claim the \mb{} method is based on -- that we can meaningfully bootstrap
a template library from a small set of labelled atlas images -- is validated in
the cross validation conducted in Experiment 1 (and the replication in
Experiment 5, Supplementary Materials). We find that both increasing the number
of atlases and the number of templates used improves \mb{} segmentation over and
above basic-\multiatlas segmentations using the same number of atlas images.
That is, by taking the extra step of generating a template library from a small
atlas set \mb{} is able to improve the final segmentation similarity. The
magnitude of this improvement is greatest with a small number of atlases, but
even with larger atlas libraries we have found that generating a template
library reduces the variability in segmentation precision (i.e.  \mb{} more
consistently produces high quality segmentations than does basic-\multiatlas{}
segmentation). These effects do not appear dependant on the hippocampal
segmentation protocol used.

@@

As when comparing against manual labels in Experiment 1, we find similar
behaviour when comparing \mb{} segmentations to \snt{} labels: similarity scores
increase with increasing numbers of atlases and templates, with diminishing
increases in improvement trending towards a plateau (Figure
\ref{fig:ADNI1-xval-k-mean}).  As in Experiment 1, using \ants{} registration
leads to significantly increased similarity scores, and there is no significant
difference in scores from any of the label fusion methods.  Mean DSC score peaks
at \Sexpr{snt.mb.kappa.best} when using 9 atlases, 19
templates, ANTS registration, and majority vote label fusion).  Compared to
\multiatlas segmentations, we find \mb{} segmentations show increasing
improvement with larger atlas and template libraries when using more than 9
templates and 5 or fewer atlases (Figure \ref{fig:ADNI1-SNT-xval-k-diff}). Peak
improvement (+\Sexpr{round(snt.mb.diff.best, digits=3)} DSC) is found with a
single atlas and template library of 19 images.  In addition to a mean increase
in similarity score over \multiatlas{}-based segmentation, \mb{} also shows more
consistency in similarity scores across all subjects and validation folds
(Figure \ref{ADNI1-SNT-xval-variability}) with a large enough template library.

@@

ggplot(subset(stats.tests, direction > 0 & stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=0.5) +
  geom_point(size=2) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Variability (p)") +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb{} vs. \multiatlas{} segmentation
  with respect to \snt segmentation.}
  Variance of segmentation accuracy between \mb{} and \multiatlas{} segmentation
  is computed for each subject across over 10 folds of Monte Carlo       
  cross validation.  
  Shown on the y-axis is the p-value from a t-test comparing the distribution of
  variances at each parameter setting (atlas/template library size).  Only points
  where \mb{} mean variability is lower than \multiatlas{} are shown.  Dashed
  lines indicate a p-value of 0.05 and 0.01.}
  \label{ADNI1-SNT-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

The gold standard for neuroanatomical segmentation is manual delineation by an
expert human rater.  However, with the availability of increasingly large MRI
datasets the time and expertise required for manual segmentation becomes
prohibitive \citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. This
effort is complicated by the fact that there is significant variation between
segmentation protocols with respect to specific anatomical boundaries of the
hippocampus \citep{Geuze2004} and this has led to efforts to create an unified
hippocampal segmentation protocol \citep{Jack2011,Boccardi2013,Boccardi2013a}.
In addition, there is controversy over the appropriate manual segmentation
protocol to use in a particular imaging study \citep{Nestor2012}. Thus, a
segmentation algorithm that can easily adapt to different manual segmentation
definitions would be of significant benefit to the neuroimaging community.

@@

Also of interest to this manuscript are the methods that attempt to define
hippocampal subfields using standard T1- or T2-weighted data, of which there are
few.  \citet{VanLeemput2009} demonstrate that the applicability of hippocampal
subfield segmentation in T1-weighted images by Bayesian techniques using Markov
random field shape priors learned from 10 manual segmentations.  This work,
available as part of the \freesurfer{} package, is limited as the segmentation
omits the tail of the hippocampus and the protocol has yet to be fully
validated.  \citet{Yushkevich2009} manually segment hippocampal subfields on
high-resolution (either $0.2mm^3$ isotropic or $0.2mm \times 0.3mm \times 0.2mm$
resolution voxels) T2-weighted MR images acquired from five post-mortem medial
temporal lobe samples. Then, using nonlinear registration guided by shape-based
models of the subfield segmentations, and manually derived hippocampus masks of
the target images, the authors demonstrate accurate parcellation of hippocampal
subfields in clinical 3T T1-weighted MRI volumes.  Using \multiatlas{} with bias
correction techniques, \citet{Yushkevich2010} demonstrate a semi-automated
method of subfield segmentation on in vivo focal T2-weighted MR acquisitions of
the temporal lobe. Manual input is only needed to mark mark divisions between
the head, body and tail of the hippocampus on target images.

@@

The following section describes experiments conducted to assess the
segmentation quality of the \mb{} algorithm:  
\begin{itemize}
\item Experiment 1 investigates the accuracy of \mb{} whole hippocampus
segmentation of aging and Alzhiemer's diseased subjects over a wide range of
parameter settings using a Monte Carlo cross-validation design. The results of
this experiment enable us to choose the parameter settings offering the best
performance for use in subsequent experiments.  
\item Experiment 2 consists of a similar cross-validation design to verify these
parameter settings on brain images of young, first episode psychosis
patients. 
\item Experiment 3 compares \mb{} whole hippocampus segmentations with those of
well-known automated methods on the entire \adnidataset{} image dataset consisting
of 
\item Experiment 4 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. 
\Sexpr{length(means.complete$RID)} brain images of subjects diagnosed as
cognitively normal, having mild cognitive impairment, or Alzheimer's disease.
\end{itemize}

@@

@@

\subsubsection{Materials and Methods: \expadni}
\paragraph{\adnidataset{} dataset} 
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
\snt{}, \maper, \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).

@@

The core claim the \mb{} method is based on -- that we can meaningfully bootstrap
a template library from a small set of labelled atlas images -- is validated in
the cross validation conducted in Experiment 1 (and the replication in
Experiment 5, Supplementary Materials). We find that both increasing the number
of atlases and the number of templates used improves \mb{} segmentation over and
above basic-\multiatlas{} segmentations using the same number of atlas images.
That is, by taking the extra step of generating a template library from a small
atlas set \mb{} is able to improve the final segmentation similarity. The
magnitude of this improvement is greatest with a small number of atlases, but
even with larger atlas libraries we have found that generating a template
library reduces the variability in segmentation precision (i.e.  \mb{} more
consistently produces high quality segmentations than does basic-\multiatlas{}
segmentation). These effects do not appear dependant on the hippocampal
segmentation protocol used.

@@

As when comparing against manual labels in Experiment 1, we find similar
behaviour when comparing \mb{} segmentations to \snt{} labels: similarity scores
increase with increasing numbers of atlases and templates, with diminishing
increases in improvement trending towards a plateau (Figure
\ref{fig:ADNI1-xval-k-mean}).  As in Experiment 1, using \ants{} registration
leads to significantly increased similarity scores, and there is no significant
difference in scores from any of the label fusion methods.  Mean DSC score peaks
at \Sexpr{snt.mb.kappa.best} when using 9 atlases, 19
templates, ANTS registration, and majority vote label fusion).  Compared to
\multiatlas{} segmentations, we find \mb{} segmentations show increasing
improvement with larger atlas and template libraries when using more than 9
templates and 5 or fewer atlases (Figure \ref{fig:ADNI1-SNT-xval-k-diff}). Peak
improvement (+\Sexpr{round(snt.mb.diff.best, digits=3)} DSC) is found with a
single atlas and template library of 19 images.  In addition to a mean increase
in similarity score over \multiatlas{}-based segmentation, \mb{} also shows more
consistency in similarity scores across all subjects and validation folds
(Figure \ref{ADNI1-SNT-xval-variability}) with a large enough template library.

@@

ggplot(subset(stats.tests, direction > 0 & stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=0.5) +
  geom_point(size=2) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Variability (p)") +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{\textbf{Difference in variability of \mb{} vs. \multiatlas{} segmentation
  with respect to \snt{} segmentation.}
  Variance of segmentation accuracy between \mb{} and \multiatlas{} segmentation
  is computed for each subject across over 10 folds of Monte Carlo       
  cross validation.  
  Shown on the y-axis is the p-value from a t-test comparing the distribution of
  variances at each parameter setting (atlas/template library size).  Only points
  where \mb{} mean variability is lower than \multiatlas{} are shown.  Dashed
  lines indicate a p-value of 0.05 and 0.01.}
  \label{ADNI1-SNT-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
